The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-11-01 talk/vmil2012/presentation/talk.tex
swap slides
trying to make diagrams, needs lots of steps
extremely minimalistic conclusion
tweak&comment

\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}

@@

@@

\begin{frame}
  \frametitle{Guards}
  \begin{itemize}
      \item Points of control flow divergence are marked with guards
      \item Operations that check whether conditions are still true
      \item When a guard fails, execution of the trace stops and continues in the interpreter
      \pause
      \item \emph{This talk:} technology and design decisions of guards
      \pause
      \begin{block}{Guard Characteristics}
          \begin{itemize}
              \item lots of them, up to 20\% guards
              \item most never fail
              \item need big information attached to them
          \end{itemize}
      \end{block}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{RPython and PyPy}
  \begin{itemize}
      \item Context: RPython
      \item a generic tracing JIT, applicable to many languages
      \item main use: PyPy, an efficient Python interpreter
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Symbolic Frame Capturing}
  \begin{itemize}
      \item Guard can fail deep inside inlined function
      \item when going back to the interpreter, call stack needs to be re-created
      \item done with the help of symbolic frame stacks
      \item these show how trace variables fill the to-be-built stack frames
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Symbolic Frame Compression}
  \begin{itemize}
      \item There are \emph{a lot of} guards
      \item Naively storing symbolic frames would be costly in terms of memory
      \item need to store them compactly
      \item observation: from one guard to the next, the non-top stack frames don't change
      \item share these between subsequent guards
      \pause
      \item also need a byte-saving binary representation, but that's just boring work
  \end{itemize}
\end{frame}

@@

@@

\begin{frame}
  \frametitle{Emitting Guards}
  Guards are compiled as
  \begin{itemize}
    \item quick Check if the condition holds
    \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
  \end{itemize}
  \pause
  In case of failure
  \begin{itemize}
    \item execution jumps to shared compensation code, decodes and stores mapping
    \item returns to interpreter that rebuilds state
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Compiling a Bridge}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_compiled.pdf}
  \end{figure}
\end{frame}
\begin{frame}
  \frametitle{Patching Guards for Bridges}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_patched.pdf}
  \end{figure}
\end{frame}
\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}
\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates / Go Benchmark}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/go_data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}

%\section{Evaluation}

>>>>

\title[Guards in RPython's Tracing JIT]{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}

@@

\begin{frame}
  \frametitle{RPython and PyPy}
  \begin{itemize}
      \item Context: RPython
      \item a language for writing interpreters for dynamic languages
      \item a generic tracing JIT, applicable to many languages
      \item used to implement PyPy, an efficient Python interpreter
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{How fast is PyPy?}
  \includegraphics[scale=0.3]{figures/all_numbers.png}
\end{frame}

@@

\begin{frame}
  \frametitle{Guards}
  \begin{itemize}
      \item Points of control flow divergence are marked with guards
      \item Operations that check whether conditions are still true
      \item When a guard fails, execution of the trace stops and continues in the interpreter
      \pause
      \item similar to deoptimization points, but more common, and patchable
      \item \emph{This talk:} technology and design decisions of guards
      \pause
      \begin{block}{Guard Characteristics}
          \begin{itemize}
              \item lots of them, up to 20\% guards
              \item most never fail
              \item need big information attached to them
          \end{itemize}
      \end{block}
  \end{itemize}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.6]{figures/op_percentage_after}
\end{frame}

\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates / Go Benchmark (29989 Guards)}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/go_data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}

\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}

@@

@@

\begin{frame}
  \frametitle{Symbolic Frame Capturing}
  \begin{itemize}
      \item Guard can fail deep inside inlined function
      \item when going back to the interpreter, call stack needs to be re-created
      \item done with the help of symbolic frame stacks
      \item these show how trace variables fill the to-be-built interpreter stack frames
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Symbolic Frame Compression}
  \begin{itemize}
      \item There are \emph{a lot of} guards
      \item Naively storing symbolic frames would be costly in terms of memory
      \item need to store them compactly
      \item observation: from one guard to the next, the non-top stack frames don't change
      \item share these between subsequent guards
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Compact Representation}
    also need a byte-saving binary representation, but that's just boring work
\end{frame}

@@

\begin{frame}
  \frametitle{Emitting Guards}
  Guards are compiled as
  \begin{itemize}
    \item quick check if the condition holds
    \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
  \end{itemize}
  \pause
  In case of failure
  \begin{itemize}
    \item execution jumps to shared compensation code, decodes and stores mapping
    \item returns to interpreter that rebuilds state
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Compiling a Bridge}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_compiled.pdf}
  \end{figure}
\end{frame}
\begin{frame}
  \frametitle{Patching Guards for Bridges}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_patched.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{JIT memory overhead}
  \includegraphics[width=\textwidth]{figures/jit_memory}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.6]{figures/resume_data_size}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
      \item Things that sound simple still often need careful engineering
      \pause
      \item guards are fundamental part of tracing JITs, need to be implemented well
      \item not even any direct performance gains
      \item keep memory usage sane
      \item allows good bridges
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Thank you! Questions?}
  \begin{itemize}
      \item Things that sound simple still often need careful engineering
      \item guards are fundamental part of tracing JITs, need to be implemented well
      \item not even any direct performance gains
      \item keep memory usage sane
      \item allows good bridges
  \end{itemize}
\end{frame}
%\section{Evaluation}
